---
title: "ESS212-HW1"
author: "Yiting Yu"
date: "2024-01-15"
output:
  pdf_document: default
---

[Yiting's github repository url](https://github.com/yitingyuu/ESS212_W24_yitingyuu.git)

# Problem 2
```{r}
language <- "R"
cat(sprintf("%s says: Hello, World!\n", language))
```

# Problem 3

$$
\begin{aligned}
S_1(n)&\equiv 1+2+3+...+n=\frac{n(n+1)}{2}\\
O_1(n)&\equiv 1+3+5+...+(2n-1)=n^2\\
S_2(n)&\equiv 1^2+2^2+3^2+...+n^2=\frac{n(n+1)(n+2)}{6}
\end{aligned}
$$

```{r}
# S_1(n)
# iterative algorithm
S1_iter <- function(n) {
  S <- 0
  for (i in  1:n) {
    S <- S + i
    }
  return(S) 
  }

# recursive algorithm
S1_rec <- function(n) {
  if (n > 1) {
    S = n + S1_rec(n-1)
    }
  else{
    S = 1
    }
  return(S) 
  }
```

### Verify
```{r}
n <- c(1, 10, 100)
(r_s1 <- data.frame(
  n = n,
  iter_result = sapply(n, S1_iter),
  rec_result = sapply(n, S1_rec),
  real_result = n*(n+1)/2
))
```

* Describe the internal state of the programs during the computational process. 

* Using either the recursive algorithm or the iterative algorithm, write and test computer programs to evaluate O1(n) and S2(n).

```{r}
# O_1(n)
# iterative algorithm
O1_iter <- function(n) {
  S <- 0
  for (i in 1:n) {
    S <- S + 2*i-1
  }
  return(S)
}

# recursive algorithm
O1_rec <- function(n) {
  if (n > 1) {
    S = 2*n-1 + O1_rec(n-1)
  } else {
    S = 1
  }
}

# verify
n <- c(1, 10, 100)
(r_o1 <- data.frame(
  n = n,
  iter_result = sapply(n, O1_iter),
  rec_result = sapply(n, O1_rec),
  real_result = n^2
))
```

```{r}
# S_2(n)
## iterative algorithm
S2_iter <- function(n) {
  S <- 0
  for (i in 1:n) {
    S <- S + i^2
  }
  return(S)
}
## recursive algorithm
S2_rec <- function(n) {
  if (n > 1) {
    S = n^2 + S2_rec(n-1)
    }
  else{
    S = 1
    }
  return(S) 
  }

## verify
n <- c(1, 10, 100)
(r_s2 <- data.frame(
  n = n,
  iter_result = sapply(n, S2_iter),
  rec_result = sapply(n, S2_rec),
  real_result = (n*(n+1)*(2*n+1))/6
))
```

## Problem 4

$$
\begin{aligned}
C(n,k)=C(n-1,k-1)+C(n-1,k)
\end{aligned}
$$

* Write a procedure in your favorite computer language that uses Equation (7) evaluates $C(n, k)$ using a recursive algorithm.

```{r}

```

* Write a procedure in your favorite computer language that takes an integer n as input and prints out the n-th row of Pascal’s triangle.

```{r}

```