---
title: "ESS212-HW1"
author: "Yiting Yu"
date: "2024-01-15"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# Problem 1
[Yiting's Github repository url](https://github.com/yitingyuu/ESS212_W24_yitingyuu.git)

# Problem 2
```{r}
language <- "R"
cat(sprintf("%s says: Hello, World!\n", language))
```

Please see the GitHub repository for Python, Julia, and Matlab code for question 2.

# Problem 3

$$
\begin{aligned}
S_1(n)&\equiv 1+2+3+...+n=\frac{n(n+1)}{2}\\
O_1(n)&\equiv 1+3+5+...+(2n-1)=n^2\\
S_2(n)&\equiv 1^2+2^2+3^2+...+n^2=\frac{n(n+1)(n+2)}{6}
\end{aligned}
$$

```{r}
# S_1(n)
# iterative algorithm
S1_iter <- function(n) {
  S <- 0
  for (i in  1:n) {
    S <- S + i
    }
  return(S) 
  }

# recursive algorithm
S1_rec <- function(n) {
  if (n > 1) {
    S = n + S1_rec(n-1)
    }
  else{
    S = 1
    }
  return(S) 
  }
```

### Verify
```{r}
n <- c(1, 10, 100)
(r_s1 <- data.frame(
  n = n,
  iter_result = sapply(n, S1_iter),
  rec_result = sapply(n, S1_rec),
  real_result = n*(n+1)/2
))
```

### Describe the internal state of the programs during the computational process. 
 
The iterative algorithm initializes a sum variable S to 0 and iterates from 1 to n, accumulating the sum of numbers in each iteration. Once it reaches n, the function returns S, the total sum of the first n positive integers.   
The recursive algorithm for summing integers from 1 to n decrements n and accumulates values until n is 1. As recursion unwinds, it adds each n to the total, returning the sum of integers from 1 to the original n.

### Using either the recursive algorithm or the iterative algorithm, write and test computer programs to evaluate O1(n) and S2(n).

```{r}
# O_1(n)
# iterative algorithm
O1_iter <- function(n) {
  S <- 0
  for (i in 1:n) {
    S <- S + 2*i-1
  }
  return(S)
}

# recursive algorithm
O1_rec <- function(n) {
  if (n > 1) {
    S = 2*n-1 + O1_rec(n-1)
  } else {
    S = 1
  }
}

# verify
n <- c(1, 10, 100)
(r_o1 <- data.frame(
  n = n,
  iter_result = sapply(n, O1_iter),
  rec_result = sapply(n, O1_rec),
  real_result = n^2
))
```

```{r}
# S_2(n)
## iterative algorithm
S2_iter <- function(n) {
  S <- 0
  for (i in 1:n) {
    S <- S + i^2
  }
  return(S)
}
## recursive algorithm
S2_rec <- function(n) {
  if (n > 1) {
    S = n^2 + S2_rec(n-1)
    }
  else{
    S = 1
    }
  return(S) 
  }

## verify
n <- c(1, 10, 100)
(r_s2 <- data.frame(
  n = n,
  iter_result = sapply(n, S2_iter),
  rec_result = sapply(n, S2_rec),
  real_result = (n*(n+1)*(2*n+1))/6
))
```

Please see the GitHub repository for Python, Julia, and Matlab code for question 3.

# Problem 4

$$
\begin{aligned}
C(n,k)=C(n-1,k-1)+C(n-1,k)
\end{aligned}
$$

### Write a procedure in your favorite computer language that uses Equation (7) evaluates $C(n, k)$ using a recursive algorithm.

```{r}
bino_coef <- function(n,k){
  if (k < n && k > 0) {
    C = choose(n-1, k-1) + choose(n-1, k)
  }
  else{
    if (k > n){
      C = 0
    }
    if (k == 0 || k == n){
      C = 1
    }
  }
  return(C)
}

# verify 
n <- 180
k <- 23
(r_bino_coef <- data.frame(
  n = n, k = k,
  r1 = bino_coef(n, k),
  r2 = choose(n, k)
))
```

### Write a procedure in your favorite computer language that takes an integer n as input and prints out the n-th row of Pascal’s triangle.

```{r}
Pt_nrow <- function(n) {
  row <- numeric(2*n-1)
  row[seq(1, 2*n-1, by=2)] <- sapply(0:(n-1), function(k) choose(n-1, k))
  
  # Print the row
  cat(row, "\n")
}

# verify:
Pt_nrow(9)
Pt_nrow(10)
```

# Problem 5

$$
\begin{aligned}
S_n = \frac{a(1-r^n)}{1-r}
\end{aligned}
$$

```{r}
S_n <- function(a, n, r) {
  if (n > 1) {
    Sn = a*(r^(n-1)+S_n(a,n-1,r))
  } 
  else {
    Sn = a
  }
  return(Sn)
}

```

### Write unit tests to demonstrate that your function works. Your tests should include cases where $r<0, r=0, 0<r<1, r=1, and \,r>1$.
```{r}
library(testthat)
a <- 1
n <- 10
# unit tests
test_that("Test S_n function with various r values", {
  # r < 0
  expect_equal(S_n(a, n, -1), a*(1-(-1)^n) / (1-(-1)))
  # r = 0
  expect_equal(S_n(a, n, 0), a)
  # 0 < r < 1
  expect_equal(S_n(a, n, 0.5), a*(1-(0.5)^n) / (1-0.5))
  # r = 1
  expect_equal(S_n(a, n, 1), a*n)
  # r > 1
  expect_equal(S_n(a, n, 1.5), a*(1-1.5^n) / (1-1.5))
})

```

### For what values of r does the limit $\lim_{n\to\infty}$ $S_n$ exist? What is the limiting value?

$$
\begin{aligned}
S_n = \frac{a(1-r^n)}{1-r}
\end{aligned}
$$

When $|r|<1$, the limit $\lim_{n\to\infty}$ $S_n$ exist, the limiting value is $\frac{a}{1-r}$.   
When $|r|\geq1$, the limit $\lim_{n\to\infty}$ $S_n$ doesn't exist. 
