---
title: "ESS212-HW2"
author: "Yiting Yu"
date: "2024-02-09"
output: pdf_document
---

[Yiting's Github repository url](https://github.com/yitingyuu/ESS212_W24_yitingyuu.git)

## Problem 1

$$
\begin{aligned}
L[n+2]=&L[n+1]+L[n]\\
L[0]=&2\\
L[1]=&1
\end{aligned}
$$

### a)

$$
\begin{aligned}
L[2]=L[0]+L[1]=3\\
L[3]=L[2]+L[1]=4\\
L[4]=L[3]+L[2]=7
\end{aligned}
$$

### b)

$$
\begin{aligned}
L[n]=&A(\frac{1+\sqrt 5}{2})^n-B(\frac{1-\sqrt 5}{2})^n\\
L[0]=&A(\frac{1+\sqrt 5}{2})^0-B(\frac{1-\sqrt 5}{2})^0\\
=&A-B\\
\Rightarrow& A-B=2\\
L[1]=&A\frac{1+\sqrt 5}{2}-B\frac{1-\sqrt 5}{2}=1\\
\Rightarrow&
\begin{cases}
A-B=2\\
A\frac{1+\sqrt 5}{2}-B\frac{1-\sqrt 5}{2}=1
\end{cases}
\begin{cases}
A=1\\
B=-1
\end{cases}\\
verify:L[n]=&(\frac{1+\sqrt 5}{2})^n+(\frac{1-\sqrt 5}{2})^n\\
L[n+1]=&(\frac{1+\sqrt 5}{2})^{n+1}+(\frac{1-\sqrt 5}{2})^{n+1}\\
L[n]+L[n+1]=&(\frac{1+\sqrt 5}{2})^n(1+\frac{1+\sqrt 5}{2})+(\frac{1-\sqrt 5}{2})^n(1+\frac{1-\sqrt 5}{2})\\
=&(\frac{1+\sqrt 5}{2})^n\frac{3+\sqrt 5}{2}+(\frac{1-\sqrt 5}{2})^n\frac{3-\sqrt 5}{2}\\
=&(\frac{1+\sqrt 5}{2})^{n+2}+(\frac{1-\sqrt 5}{2})^{n+2}=L(n+2)
\end{aligned}
$$

## Problem 2

### a)

![Figure 1: Block diagram for an input output system Problem 2(a)](/Users/m.f./Desktop/24win/ess212/homework/figure 1)

$$ 
\begin{aligned}
y[n]=\alpha(y[n-1]+x[n])
\end{aligned}
$$

### b)

![Figure 2: Block diagram for an input output system Problem 2(b)](/Users/m.f./Desktop/24win/ess212/homework/figure 2)

$$ 
\begin{aligned}
y[n]=\alpha(x[n]+x[n-1]+y[n-2])
\end{aligned}
$$

## Problem 3

### a)

$$ 
\begin{aligned}
y[n]=0.2y[n-1]+\beta x[n]
\end{aligned}
$$

![Figure 3: Block diagram for an input output system Problem 3(a)](/Users/m.f./Desktop/24win/ess212/homework/figure 3)

### b)

$$ 
\begin{aligned}
y[n]=2y[n-1]-y[n-2]+x[n]+x[n-1]
\end{aligned}
$$

![Figure 4: Block diagram for an input output system Problem 3(b)](/Users/m.f./Desktop/24win/ess212/homework/figure 4)

## Problem 4

### a)

$$ 
\begin{aligned}
y[n]=&0.2y[n-1]+\beta x[n]\\
y[1]=&0.2y[0]+\beta x[1]\\
y[0]=&0.2y[-1]+\beta x[0]
\end{aligned}
$$

No, we need to know the value of $y[-1]$ to compute the value of $y[n]$ for $n=0,1,2,3,...,N$.

### b)
If $x[n]=0$ for $n>N$, then $y[n]=0.2y[n-1]$ for $n>N$.

## Problem 5

$$
\begin{aligned}
S[n+2]=&2S[n+1]+2S[n]\\
S[0]=&0\\
S[1]=&1
\end{aligned}
$$

### a)

```{r}
# a)
fn_Sn <- function(N) {
  S <- numeric(N + 1)  # initialize S with enough space
  
  # initial conditions
  S[1] <- 0  # S[0] = 0
  S[2] <- 1  # S[1] = 1
  
  if (N == 0) {
    return(S[1])  # return S[0] for N = 0
  }
  
  for (n in 1:N) {
    S[n + 2] <- 2 * S[n + 1] + 2 * S[n]  
  }
  
  return(S[N + 1])  # return the N-th term
}
```

### b)

$$
\begin{aligned}
S[n]=Aa^n+Bb^n
\end{aligned}
$$

$$
\begin{aligned}
S[n+2]=&2S[n+1]+2S[n]\\
\Rightarrow &r^2=2r+2\\
\Rightarrow &r^2-2r-2=0\\
&r=\frac{2\pm\sqrt{(-2)^2-4\times (-2)}}{2}\\
&a=1+\sqrt{3}\\
&b=1-\sqrt{3}\\
\end{aligned}
$$

$$
\begin{aligned}
initial\, condition:&
\begin{cases}
S[0]=0\\
S[1]=1\\
\end{cases}\\
&\Rightarrow 
\begin{cases}
A+B=0\\
(1+\sqrt3)A+(1-\sqrt3)B=1
\end{cases}\\
&\Rightarrow
\begin{cases}
A=\frac{\sqrt{3}}{6}\\
B=-\frac{\sqrt{3}}{6}
\end{cases}\\
&\Rightarrow
S[n]=\frac{\sqrt{3}}{6}(1+\sqrt{3})^n-\frac{\sqrt{3}}{6}(1-\sqrt{3})^n
\end{aligned}
$$

### c)
```{r}
compute_Sn <- function(n) {
  A <- sqrt(3) / 6
  B <- - sqrt(3) / 6
  a <- 1 + sqrt(3)
  b <- 1 - sqrt(3)
  Sn <- A * a^n + B * b^n  
  return(Sn)
}

test_fn_Sn <- function() {
  passed <- TRUE
  tol <- 1e-6  
  
  for (n in 0:20) {
    expected <- compute_Sn(n)
    actual <- fn_Sn(n)
    
    if (abs(expected - actual) > tol) {
      cat(sprintf("Test failed for n = %d: expected %f, got %f\n", n, expected, actual))
      passed <- FALSE
    } else {
      cat(sprintf("Test passed for n = %d: expected %f, got %f\n", n, expected, actual))
    }
  }
  
  if (passed) {
    cat("All tests passed.\n")
  } else {
    cat("Some tests failed.\n")
  }
}

test_fn_Sn()
```

## Problem 6

### a)

$$
\begin{aligned}
x[n+1]=\frac{1}{2}(x[n]+\frac{c}{x[n]})
\end{aligned}
$$

$$
\begin{aligned}
Newton's \, method:& \\
x[n+1]=&x[n]-\frac{f(x[n])}{f'(x[n])}\\
set\, f(x)=&x^2-c\\
f(x[n])=&x[n]^2-c,\, f'([n])=2x[n]\\
x[n+1]=&x[n]-\frac{x[n]^2-c}{2x[n]}\\
=&x[n]-\frac{1}{2}x[n]+\frac{c}{2x[n]}\\
=&\frac{1}{2}(x[n]+\frac{c}{x[n]})
\end{aligned}
$$

Therefore, the above nonlinear recurrence relation can be used to compute the square root of c.

### b)

```{r}
sqrt_est <- function(c, tol=1e-6) {
  if (c < 0) {
    stop("c must be non-negative")
  }
  
  # initial guess for the square root of c
  x <- ifelse(c > 0, c / 2, 0)
  
  while (abs(x^2 - c) >= tol) {
    x <- 0.5 * (x + c / x)
  }
  
  return(x)
}

# e.g.
(sqrt_est(1225))
(sqrt_est(pi))
(sqrt_est(0.1004))
```

### c)

```{r}
library(testthat)

test_that("sqrt_est correctly estimates the square root", {
  expect_equal(sqrt_est(1225), sqrt(1225), tol = 1e-6)
  expect_equal(sqrt_est(pi), sqrt(pi), tol = 1e-6)
  expect_equal(sqrt_est(0.1004), sqrt(0.1004), tol = 1e-6)
})
```

### d)
The values of $x_0=x[0]$ should be positive to converge the recurrence relation to the square root of c.

## Problem 7

```{r fig.width=6, fig.height=6}
library(ggplot2)

mandelbrot <- function(c, max_iter) {
  z <- 0
  for(i in 1:max_iter) {
    z <- z^2 + c
    if(Mod(z) > 2) {
      return(FALSE)
    }
  }
  return(TRUE)
}

x_min <- -2
x_max <- 0.5
y_min <- -1.5
y_max <- 1.5
x <- seq(x_min, x_max, by = 5e-3)
y <- seq(y_min, y_max, by = 5e-3)
grid <- expand.grid(x = x, y = y)

# check each point in the grid
max_iter <- 500
points <- vector("list", length=nrow(grid))
for(i in 1:nrow(grid)) {
  c <- complex(real = grid$x[i], imaginary = grid$y[i])
  points[[i]] <- mandelbrot(c, max_iter)
}

# convert the results to a data frame
grid$set <- unlist(points)

ggplot(grid, aes(x, y)) + 
  geom_tile(aes(fill = set)) + 
  scale_fill_manual(values = c("TRUE" = "snow", "FALSE" = "darkslateblue")) +
  scale_y_continuous(breaks = seq(-1.5, 1.5, by = 0.5)) +  theme_minimal() + 
  labs(title = "Mandelbrot Set", x = "Real(c)", y = "Imag(c)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

## Problem 8

$$
\begin{aligned}
E_0=\frac{1}{2}mv^2+mgz\\
t_0=0,\, v(0)=0\\
E_o=mgz_0\, at\, time\, t_0=0\\
\Rightarrow \frac{dz}{dt}=v=-\sqrt{2g(z-z_0)}\\
z(0)=z_0\\
z(t+\Delta t)=&z(t)+\Delta tv\\
=& z(t)-\Delta t (\sqrt{2g(z-z_0)})
\end{aligned}
$$

### a)

$$
\begin{aligned}
z(\Delta t)=&z(0+\Delta t)\\
=&z(0)+\Delta tv\\
=&z_0-\Delta t (\sqrt{2g(z-z_0)})\\
z(2\Delta t)=&z(0+2\Delta t)\\
=&z(0)+2\Delta tv\\
=&z_0-2\Delta t (\sqrt{2g(z-z_0)})\\
\end{aligned}
$$

```{r}
g <- 9.81 
z0 <- 100 
delta_t <- 1e-6 
v <- 0 

# Will-E scheme, Euler method 
z <- z0 - 1e-6 
t <- 0 

while (z > 0) {
  v <- sqrt(2 * g * (z0 - z)) 
  delta_z <- v * delta_t
  if (z - delta_z < 0) {
    z <- 0
  } else {
    z <- z - delta_z # Loop until z <= 0
  }
  t <- t + delta_t 
}

real_t <- sqrt(2 * z0 / g)

# Results
cat("Time when z reaches 0 calculated by the Will-E scheme:", t, "seconds\n")
cat("Time when z reaches 0 by the physical fact:", real_t, "seconds\n")
```

The expected time for Will-E to reach the bottom of the canyon is 4.51479 seconds. Will-E's scheme requires a very small time step ($\Delta t$), or the time computed using the Euler method is inaccurate. However, choosing a small enough time step may result in computationally inefficient results. Will-E's scheme doesn't satisfy the law of conservation of energy. 

## b)
$$
\begin{aligned}
m\frac{d^2z}{dt^2}=&-mg\\
\frac{dz}{dt}=&v\\
\frac{dv}{dt}=&-g
\end{aligned}
$$

```{r}
g <- 9.81 
z0 <- 100 
m <- 1 
delta_t <- 1e-3

# initial conditions
z <- z0
v <- 0 
t <- 0

ts <- c()
ki_es <- c()
po_es <- c()
tt_es <- c()
posi <- c() 

# Euler forward method
while (z > 0) {
  v <- v + g * delta_t
  z <- max(0, z - v * delta_t) # ensure z >= 0
  t <- t + delta_t 
  
  ki_e = 0.5 * m * v^2
  po_e = m * g * z
  tt_e = ki_e + po_e
  
  ts <- append(ts, t)
  ki_es <- append(ki_es, ki_e)
  po_es <- append(po_es, po_e)
  tt_es <- append(tt_es, tt_e)
  posi <- append(posi, z) 
}

plot(ts, ki_es, type = 'l', col = 'darkblue', xlim = c(0, 5),
     ylim = c(0, max(tt_es)), xlab = "Time (s)", ylab = "Energy (J)", 
     main = "Energies vs Time")
lines(ts, po_es, col = 'darkred')
lines(ts, tt_es, col = 'darkgreen')
legend("left", inset=.05, legend = c("kinetic energy", "potential energy", "total energy"), 
       col = c("darkblue", "darkred", "darkgreen"), lty = 1, cex = 0.8)

# z = 0
bottom_t <- ts[which.min(posi > 0)]
abline(v = bottom_t, col = "black", lty = 2) 
text(bottom_t, max(tt_es) * 0.5, labels = paste("z=0 at", round(bottom_t, 6), "s"), 
     pos = 4, cex = 0.6)

print(paste("Time when z=0 calculated by the Euler method:", round(bottom_t, 6), "seconds"))
cat("Time when z reaches 0 by the physical fact:", real_t, "seconds\n")
```

The time when z reaches 0 calculated by the Euler forward scheme is better than the result computed by the Will-E scheme since it is closer to the real-time when z=0. However, the Euler forward scheme also doesn't satisfy the energy conservation principle.

## c)
```{r}
g <- 9.81 
z0 <- 100 
m <- 1 
delta_t <- 1e-3 
tt_time <- sqrt(2 * z0 / g) 
N <- as.integer(tt_time / delta_t)

v <- rep(0, N) 
z <- rep(0, N) 
times <- seq(0, tt_time, length.out = N)

# initial conditions
v[1] <- 0 
z[1] <- z0 

ki_es <- rep(0, N)
po_es <- rep(0, N)
tt_es <- rep(0, N)

# Leap-frog scheme for i=2 
v[2] <- v[1] + g * delta_t
z[2] <- z[1] - 0.5 * g * delta_t^2 

# Leap-frog iteration for i = 3 to N
for (i in 3:N) {
  v[i] <- v[i-2] + 2 * g * delta_t 
  z[i] <- z[i-2] - 2 * v[i-1] * delta_t 
  
  ki_es[i] <- 0.5 * m * v[i]^2
  po_es[i] <- m * g * (z[i] - 0) 
  tt_es[i] <- ki_es[i] + po_es[i]
}

ki_es[1] <- 0.5 * m * v[1]^2
po_es[1] <- m * g * (z[1] - 0)
tt_es[1] <- ki_es[1] + po_es[1]

plot(times[-2], ki_es[-2], type = 'l', col = 'darkblue', xlim = c(0, 5), 
     ylim = c(min(po_es[-2]), max(tt_es[-2])), xlab = "Time (s)", 
     ylab = "Energy (J)", main = "Energies vs Time") 
lines(times[-2], po_es[-2], col = 'darkred')
lines(times[-2], tt_es[-2], col = 'darkgreen')
legend("left", legend = c("kinetic energy", "potential energy", "total energy"), 
       col = c("darkblue", "darkred", "darkgreen"), lty = 1, cex = 0.8) 

t_bottom <- times[which.min(posi > 0)]
abline(v = t_bottom, col = "black", lty = 2) 
text(t_bottom, max(tt_es) * 0.5, labels = paste("z=0 at", round(t_bottom, 6), "s"), 
     pos = 4, cex = 0.6)

print(paste("Time when z=0 calculated by the leap-frog scheme:", round(t_bottom, 6), "seconds"))
cat("Time when z reaches 0 by the physical fact:", tt_time, "seconds\n")
```

The expected time for Will-E to reach the bottom computed by the leap-frog scheme is 4.515236 seconds, the same as the time calculated by the physical fact. Therefore, the leap-frog scheme satisfies the principle of conservation of energy, and it is better than the Euler forward and Will-E scheme.