---
title: "ESS212 HW3"
author: "Yiting Yu"
date: "2024-02-24"
output: pdf_document
---

[Yiting's Github repository url](https://github.com/yitingyuu/ESS212_W24_yitingyuu.git)

```{r, echo = FALSE}
library(MASS) 
library(numDeriv) 
library(httr)
library(ggplot2)
library(readr)
```

## Problem 1

$$
\begin{aligned}
(x_1,y_1)=&(-4,30)\\
(x_2,y_2)=&(0,2)\\
(x_3,y_3)=&(4,6)\\
y_i=&a(x_i)^2+bx_i+c_i+e_i&i=1,2,3\\
\Rightarrow&
\begin{cases}
30=a(-4)^2+b(-4)+c\\
2=a(0)^2+b(0)+c\\
6=a(4)^2+4b+c
\end{cases}\\
\Rightarrow&
\begin{cases}
30=16a-4b+c\\
2=c\\
6=16a+4b+c
\end{cases}\\
\Rightarrow&
\begin{cases}
a=1\\
b=-3\\
c=2
\end{cases}\\
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{d}=&\begin{bmatrix} y_1 & y_2 & y_3 \end{bmatrix}^\prime\\
\mathbf{a}=&\begin{bmatrix} x_1^2 & x_2^2 & x_3^2 \end{bmatrix}^\prime\\
\mathbf{b}=&\begin{bmatrix} x_1 & x_2 & x_3 \end{bmatrix}^\prime\\
\mathbf{e}=&\begin{bmatrix} 1 & 1 & 1 \end{bmatrix}^\prime\\
\mathbf{A}=& \begin{bmatrix}
x_1^2 & x_1 & c_1 \\
x_2^2 & x_2 & c_2 \\
x_3^2 & x_3 & c_3 \\
\end{bmatrix} = \begin{bmatrix}
16 & -4 & 1 \\
0 & 0 & 1 \\
16 & 4 & 1 \\
\end{bmatrix}\\
\mathbf{d}=& \begin{bmatrix}
30 \\
2 \\
6 \\
\end{bmatrix}\\
\mathbf{A}\mathbf{\beta}=&\mathbf{d}\\
\mathbf{\beta}=&(\mathbf{A^\prime A})^{-1}\mathbf{A}^{\prime}\mathbf{d}\\
\mathbf{e}=&\mathbf{d}-\hat{\mathbf{d}}
\end{aligned}
$$

```{r}
A <- matrix(c(16, -4, 1, 0, 0, 1, 16, 4, 1), 3, byrow = TRUE)
d <- c(30, 2, 6)

beta <- solve(A, d)
#beta <- solve(t(A) %*% A) %*% t(A) %*% d

d_hat <- A %*% beta
e <- d - d_hat

# result
beta
e
```

The above results show that:
$$
\begin{aligned}
\mathbf{\beta}=& \begin{bmatrix}
a \\
b \\
c \\
\end{bmatrix}= \begin{bmatrix}
1 \\
-3 \\
2 \\
\end{bmatrix}\\
\mathbf{e}=& \begin{bmatrix}
0 \\
0 \\
0 \\
\end{bmatrix}
\end{aligned}
$$

$$
\begin{aligned}
y_1 &= A(x_1-B)(x_1-C) \\
y_2 &= A(x_2-B)(x_2-C) \\
y_3 &= A(x_3-B)(x_3-C) \\
\end{aligned}
$$

```{r}
fn <- function(beta, x, y) {
  A <- beta[1]
  B <- beta[2]
  C <- beta[3]

  F <- numeric(length(y))
  
  for(i in 1:length(y)) {
    F[i] <- A*(x[i]-B)*(x[i]-C) - y[i]
  }
  return(F)
}

# A, B, C for initial iterate
beta <- c(A=1, B=2, C=3)

x <- c(-4, 0, 4) 
y <- c(30, 2, 6) 

# Newton's method 
n <- 100
for (iter in 1:n) {
  F <- fn(beta, x, y)
  J <- jacobian(func = function(b) fn(b, x, y), x = beta)
  delta <- solve(J) %*% F
  beta <- beta - delta
  if (max(abs(delta)) <  1e-6) {
    cat("Convergence achieved after", iter, "iterations.\n")
    break
  }
}

# result
cat("Estimates for A, B, C:\n")
print(beta)
```

The above results show that:
$$
\begin{aligned}
\Rightarrow
\begin{cases}A=1\\B=2\\C=1
\end{cases}
\end{aligned}
$$

If we choose $A=0$ as the initial iteration, the model will not work because $A=0$ will result in any value assigned to $x_i$, $y_i$ always equal to 0.That's why I rule out the value of $A=0$ as a possible value.

```{r}
B_range <- seq(0, 5, length.out = 200)
C_range <- seq(0, 5, length.out = 200)

x <- c(-4, 0, 4)
y <- c(30, 2, 6)

fn_F <- function(B, C, x, y) {
  A <- 1
  F <- numeric(length(x))
  for (i in 1:length(x)) {
    F[i] <- A*(x[i]-B)*(x[i]-C) - y[i]
  }
  return(F)
}

# initialize matrix to store convergence results
con_grid <- matrix(0, length(B_range), length(C_range))

# iterate over the ranges of B and C
for (i in 1:length(B_range)) {
  for (j in 1:length(C_range)) {
    F <- fn_F(B_range[i], C_range[j], x, y)
    con_criteria <- sum(abs(F))
    con_grid[i, j] <- ifelse(con_criteria < 1, 1, 0)
  }
}

colors <- colorRampPalette(c("lightblue", "aliceblue", "white"))(n = 2)  

image(x = B_range, y = C_range, z = con_grid, col = colors, xlab = "B", ylab = "C", 
      main = "Convergence Map")
legend("topright", legend = c("Not Converged", "Converged"), fill = c(colors[1], colors[2]))
```

## Problem 2

```{r}
# download the GISTEMP dataset
url <- "https://data.giss.nasa.gov/gistemp/tabledata_v4/GLB.Ts+dSST.txt"
save_f <- tempfile()
GET(url, write_disk(save_f))

# clean the dataset
lines <- read_lines(save_f)

# filter lines to include only those with numeric data at the beginning (years)
filter_l <- grep("^\\s*\\d{4}", lines, value = TRUE)

# determine the maximum number of columns
max_cols <- max(sapply(filter_l, function(line) length(strsplit(line, "\\s+")[[1]])))

col_names <- c("Year", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", 
               "Oct", "Nov", "Dec", "J-D", "D-N", "DJF", "MAM", "JJA", "SON")
extra_cols <- max_cols - length(col_names)
if(extra_cols > 0) {
  col_names <- c(col_names, paste0("Extra", 1:extra_cols))
}

# combine the filtered lines into a single character vector
data_text <- paste(filter_l, collapse = "\n")

# read the combined text as a table
data <- read.table(text = data_text, header = FALSE, fill = TRUE, col.names = col_names)
clean_d <- data[data$`J.D` != "****", ]

#-----------------------------------------

# fit the model
yrs <- as.numeric(clean_d$Year)
mid_point <- mean(yrs)
annual_means <- as.numeric(clean_d$`J.D`) # `annual_means` is the mean of the annual temperature

model <- lm(annual_means ~ I(yrs - mid_point), na.action = na.exclude)

mu <- coef(model)[1] # intercept
m <- coef(model)[2]

# plotting
trend <- data.frame(Year = yrs, Trend = mu + m * (yrs - mid_point))
ggplot(data = clean_d, aes(x = Year, y = annual_means)) +
  geom_point() +
  geom_line(data = trend, aes(x = Year, y = Trend), color = "red") +
  labs(title = "Annual global temperature anomalies and trend",
       x = "Year", y = "Global Land-Ocean temperature anomaly(0.01°C)") +
  theme_minimal()

# the estimated values of mu and m
cat("Estimated mu:", mu, "\n")
cat("Estimated m:", m, "\n")

# the standard deviation of the residuals
sd_residual <- sd(resid(model))
cat("Standard deviation of the residuals:", sd_residual, "\n")
```

The above results show that:
$$
\begin{aligned}
\mu=&6.923611\\
m=&0.7868961 \\
\end{aligned}
$$

The standard deviation of the components of the $\mathbf{e}$ vector is 18.42912. 

## Problem 3
There are 16 seedlings to be planted in 4 different conditions house. Since four seedlings are planted in each greenhouse, I assume seedlings $1,5,9,13$ were planted in $Ba$ greenhouse, $2,6,10,14$ were planted in $Bb$ greenhouse, $3,7,11,15$ were planted in $Aa$ greenhouse, and seedlings $4,8,12,16$ were planted in $Ab$ greenhouse.

$$
\begin{aligned}
d_i=&\mu+\beta_1\Delta_{CO_2,i}+\beta_2\Delta_{H_2O,i}+\beta_3\Delta_{CO_2,i}\Delta_{H_2O,i}+e_i\\
\Rightarrow&
\begin{bmatrix}
d_1 \\d_2 \\d_3 \\d_4 \\
\end{bmatrix}=
\begin{bmatrix}
d_5 \\d_6 \\d_7 \\d_8 \\
\end{bmatrix}=
\begin{bmatrix}
d_9 \\d_{10} \\d_{11} \\d_{12} \\
\end{bmatrix}=
\begin{bmatrix}
d_{13} \\d_{14} \\d_{15} \\d_{16} \\
\end{bmatrix}=
\begin{bmatrix}
1 & 1 & 0 & 0 \\
1 & 1 & 1 & 1 \\
1 & 0 & 0 & 0 \\
1 & 0 & 1 & 0 \\
\end{bmatrix}\\
\Rightarrow
\mathbf{d}=&\mathbf{A\beta}+\mathbf{e}\\
\Rightarrow
\mathbf{d}=&\begin{bmatrix}
d_1 \\d_2 \\d_3 \\d_4 \\
d_5 \\d_6 \\d_7 \\d_8 \\
d_9 \\d_{10} \\d_{11} \\d_{12} \\
d_{13} \\d_{14} \\d_{15} \\d_{16} \\
\end{bmatrix}
\mathbf{A}=
\begin{bmatrix}
1 & 1 & 0 & 0 \\1 & 1 & 1 & 1 \\1 & 0 & 0 & 0 \\1 & 0 & 1 & 0 \\
1 & 1 & 0 & 0 \\1 & 1 & 1 & 1 \\1 & 0 & 0 & 0 \\1 & 0 & 1 & 0 \\
1 & 1 & 0 & 0 \\1 & 1 & 1 & 1 \\1 & 0 & 0 & 0 \\1 & 0 & 1 & 0 \\
1 & 1 & 0 & 0 \\1 & 1 & 1 & 1 \\1 & 0 & 0 & 0 \\1 & 0 & 1 & 0 \\
\end{bmatrix}
\mathbf{\beta}=\begin{bmatrix}
\mu\\ \beta_1\\ \beta_2\\ \beta_3
\end{bmatrix}
\end{aligned}
$$

```{r}
A <- matrix(c(
  1, 1, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0,
  1, 1, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0,
  1, 1, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0,
  1, 1, 0, 0, 1, 1, 1, 1, 1, 0, 0, 0, 1, 0, 1, 0
), nrow = 16, byrow = TRUE)

(Aprime_A <- t(A) %*% A)
```

$$
\begin{aligned}
\mathbf{A'A}=&
\begin{bmatrix}
16 & 8 & 8 & 4 \\
8 & 8 & 4 & 4 \\
8 & 4 & 8 & 4 \\
4 & 4 & 4 & 4 \\
\end{bmatrix} \\
\mathbf{A'd}=&
\begin{bmatrix}
\Sigma_{i=1}^{16} d_i \\
d_1+d_2+d_5+d_6+d_9+d_{10}+d_{13}+d_{14}\\
d_2+d_4+d_6+d_8+d_{10}+d_{12}+d_{14}+d_{16} \\
d_2+d_6+d_{10}+d_{14} \\
\end{bmatrix} \\
\end{aligned}
$$